---
title: "College Affordablity"
author: "Alden Cowap"
date: "12/10/2016"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project aims to create a useful tool for students applying to college by allowing them to find schools they will be able to afford. Students are able to select their academic crudentials, household income level, and the most they are willing to pay for tuition and then see an interactive map that includes all of the schools that fit their needs. The hope is that this will allow students to move beyond sticker shock at the listed prices of colleges and universities and see all of their possiblities depending on their limitations. This interactive map is proceeded by a brief examination of the realtionship between college affordability and college endowment across the United States. All of the data below comes from 2012-2013.  

```{r, echo=FALSE, message=FALSE, warning=FALSE}

## Get data and clean data

library(stringr)
library(tm)
library(tidyverse)
library(USAboundaries)
library(maptools)
library(sp)
library(broom)
library(leaflet)
library(stargazer)
library(ggthemes)


datavar <- read.csv("variables.csv")
data2013 <- read.csv("MERGED2012_13_PP.csv")
needblind <- read.csv("needblindschools.csv")

endowment <- read.csv("endowment2013.csv")

endowment2 <- endowment %>% mutate(stringlabel = as.character(Rank)) %>% 
  filter(!(Rank %in% Rank[1:12])) %>% 
  filter(!(Rank %in% Rank[42:51])) %>% 
  mutate(first = substr(stringlabel, 1, 1)) %>% 
  filter(first %in% c(0:9))

states <- as.vector(state.abb)

endowmentname <- endowment2 %>% 
  mutate(stringlabel = str_replace_all(stringlabel, "<cd>", ".")) %>% 
  mutate(namebf = removeNumbers(stringlabel)) %>% 
  mutate(namebf = removePunctuation(namebf)) %>% 
  mutate(name = removeWords(namebf, states)) 

uniname <- as.vector(endowmentname[["namebf"]])


endowmentnums <- endowmentname %>% 
  mutate(nums = str_replace_all(stringlabel,"[^0-9 , .]","")) %>% 
  mutate(nums2 = str_replace_all(stripWhitespace(nums), " ", "/")) %>% 
  mutate(nums2 = str_replace_all(nums2, "[. ,]", "")) %>% 
  mutate(nums2 = str_replace_all(nums2, "///", "/")) %>% 
  mutate(nums2 = str_replace_all(nums2, "//", "/")) %>% 
  separate(nums2, into = c("rank", "endow2013", "endow2012", "change"), sep = "/") %>% 
  mutate(name = str_replace_all(name, " ", "")) %>% 
  mutate(name = str_replace(name, "CarnegieMellonUniversityv", "CarnegieMellonUniversity")) %>% 
  mutate(name = str_replace(name, "GeorgetownUniversityDC", "GeorgetownUniversity")) %>% 
  mutate(name = str_replace(name, "LawrenceUniversityofWisconsin", "LawrenceUniversity")) %>% 
  mutate(name = str_replace(name, "NorthCarolinaATStateUniversity", "NorthCarolinaStateUniversity")) %>% 
  mutate(name = str_replace(name, "PennsylvaniaStateUniversityii", "PennsylvaniaStateUniversity")) %>% 
  mutate(name = str_replace(name, "SanJoseStateUniversityTowerFoundation", "SanJoseStateUniversity")) %>% 
  mutate(name = str_replace(name, "StJohnsCollegeix", "StJohnsCollege")) %>% 
  mutate(name = str_replace(name, "UniversityofIllinoisFoundationiv", "UniversityofIllinoisatChicago")) %>% 
  mutate(name = str_replace(name, "FloridaStateUniversityFoundation", "FloridaStateUniversity")) %>% 
  mutate(name = str_replace(name, "UniversityofNorthCarolinaatChapelHillFoundations", "UniversityofNorthCarolina")) 



data2013temp <- data2013 %>% mutate(name = str_replace_all(as.character(INSTNM), " ", "")) %>% 
  mutate(name = str_replace(name, "ColumbiaUniversityintheCityofNewYork", "ColumbiaUniversity")) %>% 
  mutate(name = str_replace(name,"CooperUnionfortheAdvancementofScienceandArt", "CooperUnion")) %>% 
  mutate(name = str_replace(name,"FairleighDickinsonUniversity-MetropolitanCampus", "FairleighDickinsonUniversity")) %>%   
  mutate(name = str_replace(name,"HolyCrossCollege", "HolyCross")) %>%   
  mutate(name = str_replace(name,"JewishTheologicalSeminaryofAmerica", "JewishTheologicalSeminary")) %>%  
  mutate(name = str_replace(name,"Lewis&ClarkCollege", "LewisClarkCollege")) %>%   
  mutate(name = str_replace(name,"MountSaintMaryCollege", "MountSaintMarysCollege")) %>% 
  mutate(name = str_replace(name,"NorthCarolinaA&TStateUniversity", "NorthCarolinaStateUniversity")) %>%  
  mutate(name = str_replace(name,"PennsylvaniaStateUniversity-MainCampus", "PennsylvaniaStateUniversity")) %>%   
  mutate(name = str_replace(name,"StJohn'sCollege", "StJohnsCollege")) %>%   
  mutate(name = str_replace(name,"TulaneUniversityofLouisiana", "TulaneUniversity")) %>%   
  mutate(name = str_replace(name,"UniversityofNewHampshire-MainCampus", "UniversityofNewHampshire")) %>%   
  mutate(name = str_replace(name,"UniversityofNorthCarolinaatChapelHill", "UniversityofNorthCarolina")) %>%   
  mutate(name = str_replace(name,"UniversityofVirginia-MainCampus", "UniversityofVirginia")) %>%   
  mutate(name = str_replace(name,"UniversityofWashington-SeattleCampus", "UniversityofWashington"))

data2013temp <- data2013temp %>% 
  mutate(NPT41_PRIV = as.numeric(as.character(NPT41_PRIV))) %>%
  mutate(NPT42_PRIV = as.numeric(as.character(NPT42_PRIV))) %>% 
  mutate(NPT43_PRIV = as.numeric(as.character(NPT43_PRIV))) %>% 
  mutate(NPT44_PRIV = as.numeric(as.character(NPT44_PRIV))) %>% 
  mutate(NPT45_PRIV = as.numeric(as.character(NPT45_PRIV))) %>% 
  mutate(NPT41_PUB = as.numeric(as.character(NPT41_PUB))) %>% 
  mutate(NPT42_PUB = as.numeric(as.character(NPT42_PUB))) %>% 
  mutate(NPT43_PUB = as.numeric(as.character(NPT43_PUB))) %>% 
  mutate(NPT44_PUB = as.numeric(as.character(NPT44_PUB))) %>% 
  mutate(NPT45_PUB = as.numeric(as.character(NPT45_PUB))) %>% 
  mutate(NPT41 = ifelse(is.na(NPT41_PRIV), NPT41_PUB, NPT41_PRIV)) %>% 
  mutate(NPT42 = ifelse(is.na(NPT42_PRIV), NPT42_PUB, NPT42_PRIV)) %>% 
  mutate(NPT43 = ifelse(is.na(NPT43_PRIV), NPT43_PUB, NPT43_PRIV)) %>% 
  mutate(NPT44 = ifelse(is.na(NPT44_PRIV), NPT44_PUB, NPT44_PRIV)) %>% 
  mutate(NPT45 = ifelse(is.na(NPT45_PRIV), NPT45_PUB, NPT45_PRIV)) %>% 
  mutate(NUM4 = as.numeric(as.character(NUM4_PRIV))) %>% 
  mutate(NUM4 = ifelse(is.na(NUM4), as.numeric(as.character(NUM4_PUB)), NUM4)) %>% 
  mutate(SCH_DEG = as.numeric(as.character(SCH_DEG)))


total2013 <- inner_join(data2013temp, endowmentnums, by = "name") %>% arrange(rank)


latlong <- read_csv("latlong.csv") %>% 
  mutate(Name = Institution) %>% 
  mutate(Name = str_replace_all(Name, " ", "")) %>% 
  mutate(Name = str_replace_all(Name, "[. , ']", "")) %>% 
  rename(Ziplat = ZIP)
  
sat <- read_csv("sat.csv") %>% select(-Institution)

char <- inner_join(latlong, sat, by="UnitID")

total2013_lat <- inner_join(total2013, char, by=c("UNITID" = "UnitID"))

allschools_lat <- inner_join(data2013temp, char, by=c("UNITID" = "UnitID"))

needblind <- needblind %>% mutate(school = str_replace_all(Name.of.School, "[^a-z A-Z]", "")) %>% 
  mutate(school = str_replace_all(school, " ", "")) %>% 
  mutate(school = str_replace(school, "MassachusettsInstituteofTechnologyMIT", "MassachusettsInstituteofTechnology")) %>% 
  mutate(school = str_replace(school, "NewYorkUniversityNYU", "NewYorkUniversity")) %>% 
  mutate(school = str_replace(school, "UniversityofSouthernCaliforniaUSC", "UniversityofSouthernCalifornia")) %>% 
  mutate(school = str_replace(school, "WashingtonUniversity", "WashingtonUniversityinStLouis")) %>% 
  mutate(school = str_replace(school, "HarvardCollege", "HarvardUniversity")) %>% 
  mutate(school = str_replace(school, "NorthCarolinaStateUniversityNCSU", "NorthCarolinaStateUniversity")) %>% 
  mutate(school = str_replace(school,"PennState", "PennsylvaniaStateUniversity")) %>% 
  mutate(school = str_replace(school,"SanJoseStateUniversitySJSU", "SanJoseStateUniversity"))
  

blindnames <- as.vector(needblind$school)

blinddata <- total2013_lat %>% filter(name %in% blindnames)

name <- as.vector(blinddata$name)

antidata <- needblind %>% filter(!(school %in% name))

```


```{r, echo=FALSE, message=FALSE, warning=FALSE}


tograph <- blinddata %>% 
  mutate(endow = as.numeric(endow2013)) %>% 
  mutate(avg_cost = as.numeric(as.character(INC_PCT_M1))) %>% 
  filter(!is.na(avg_cost))

TEMP <- tograph %>% select(endow, INSTNM, p_w_grant, NPT41, NPT42, NPT43, NPT44, NPT45)

TEMP <- TEMP %>% gather(income, amount, 4:8)


TEMP_no_nc <- TEMP %>% filter(endow!=32959)


total2013_lat <- total2013_lat %>% mutate(endow = as.numeric(as.character(endow2013)))

TEMP2 <- total2013_lat %>% select(endow, INSTNM, p_w_grant, NPT41, NPT42, NPT43, NPT44, NPT45)

TEMP2 <- TEMP2 %>% gather(income, amount, 4:8)

#incomegroups <- c("Under $30,000", "$30,000-$48,000", "$48,000-$75,000", "$75,000-$110,000", "Above $110,000")
```


The first thing I looked at was net tuition (tuition minus financial aid) against the log of endowment for all schools in my data set. The average net tuiton was reported for five different income groups, under $30,000, $30,000-$48,000, $48,000-$75,000, $75,000-$110,000, and above $110,000. 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=TEMP2, aes(x = log10(endow), y = amount, color = income)) + 
  geom_smooth(method="lm", se = FALSE) + 
  labs(title = "All Schools", x = "Log of Endowment", y = "Net Tuition") + 
  theme_minimal() + scale_colour_economist() 
```

As seen in the graph above, as the endowment of the school increases, the net tuition for students in the bottom three income groups decreases, while net tuition for those in the top two income groups increases. 


Some schools advertise themselves as being need blind, which means they do not consider the financial needs of the student when making their admission deciscion. I was curious to see if this trend woulds still apply when I looked only at schools with need blind policies. 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=TEMP_no_nc, aes(x = log10(endow), y = amount, color = income)) + 
  geom_smooth(method="lm", se = FALSE) + 
  geom_point(alpha=.4) + 
  labs(title = "Need Blind Schools", x = "Log of Endowment", y = "Net Tuition") +
  theme_minimal() + scale_colour_economist()

```

The trend of net tuition decreasing as endowment increases remains for need blind schools. 

If every school has the same number of financial aid dollars, schools that admit financial aid students will have to charge each student more. So if endowment and the percentage of students on financial aid are correlated, that could be conflating the effect we see where net tuition decreases as endowment increases. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = TEMP, aes(x=log10(endow), y=p_w_grant)) + 
  geom_smooth(method = "lm") + 
  geom_point() + 
  labs(title = "Need Blind Schools", x="Log of Endowment", y = "Percent of Students on Financial Aid") +
  theme_minimal() 

toregress <- tograph %>% select(endow, FSEND_COUNT, NPT41, NPT42, NPT43, NPT44, NPT45, NUM4, FSEND_1, FSEND_4, INC_PCT_M2, INC_PCT_M1, INC_PCT_H1, INC_PCT_H2, p_w_grant)

fitlowest <- lm(NPT41 ~ log(endow) + p_w_grant, data=tograph)


fitsecond <- lm(NPT42 ~ log(endow) + p_w_grant, data=tograph)


fitthird <- lm(NPT43 ~ log(endow) + p_w_grant, data=tograph)


fitfourth <- lm(NPT44 ~ log(endow) + p_w_grant, data=tograph)


fithigh <- lm(NPT45 ~ log(endow) + p_w_grant, data=tograph)

```

As shown above, endowment is correlated with the percentage of students on financial aid, which means the link between endowment and net tuition cost is not actually as strong as it appreared in the first two graphs I presented. 
I also wanted to check for correlation between the net tuition the students in the lowest income group and the net tuition for those in the highest income group in order to see if schools with larger endowments charged students from high income families more in order to offset the financial aid they gave to the low income students. 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = tograph, aes(x=NPT41, y=NPT45, col=log10(endow))) + geom_point() + geom_smooth(method="lm") + labs(title="Check for Correlation", x="Under $30,000", y = "Above $110,000") + theme_minimal()

fitcheck <- lm(NPT45 ~ NPT41, data=tograph)
#summary(fitcheck)

tomap <- allschools_lat %>% 
  select(NPT41, NPT42, NPT43, NPT44, NPT45, Ziplat, INSTNM, Longitude, Latitude, STABBR, UG, SCH_DEG, p_w_grant, ACT75, SAT_AVG)


tomap <- tomap %>% filter( !STABBR %in% c("AK", "HI", "PR")) %>% filter(!is.na(STABBR)) %>% 
  filter(NPT41 > 0) %>% 
  mutate(SAT_AVG = as.numeric(as.character(SAT_AVG))) %>% 
  mutate(ACTna = ifelse(is.na(ACT75), 0, ACT75)) %>% 
  mutate(SATna = ifelse(is.na(SAT_AVG), 0, SAT_AVG))
```

I did not find a strong correlation between these two. 

I ran regressions to see whether the trend lines from the first couple of graphs were statistically significant or if there were just a couple of points that were pulling the trend line one way or another. 

```{r star, results = 'asis', warning=FALSE, message=FALSE, echo=FALSE}

stargazer(fitlowest, fitsecond, fitthird, fitfourth, fithigh, title="Regression Results", align=TRUE, type = "html",
          covariate.labels=c("Log of Endowment","Percent on Aid"), 
          dep.var.labels = c("under $30","$30-48","$48-75", "$75-110", "above $110") ,
          keep.stat="n")


```


The following is the interactive map I created. Students are able to select the type of school they are looking to apply to, their academic crudentials, their household income and the maximum tution they are able to afford, and the map will display all of the schools which fit their criteria. If you click on any specific school, all of the information about the school is displayed. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
inputPanel(
  sliderInput("income", label = "Household Income:",
              min = 0, max = 200000, value = 50000, step = 500) ,
  
  sliderInput("max", label = "Maximum Annual Tutition:",
              min = 0, max = 80000, value = 30000, step = 100),
  
  selectInput("type", label = "Type of School:",
              choices = c("4 Year Undergraduate", "2 Year Undergraduate", "Technical College"), 
              selected = "4 Year Undergraduate"),
  
  numericInput("act", label = "Maximum ACT Score", value = 36),
  
  numericInput("sat", label = "Maximum SAT Score", value = 1600)
  
)

renderLeaflet({
  
  
  income <- ifelse(input$income <= 30000, 1, 
                ifelse(input$income <=48000 & input$income > 30000, 2, 
                       ifelse(input$income> 48000 & input$income <= 75000, 3,
                              ifelse(input$income>75000 & input$income<=110000, 4, 5))))
  
  schooltype <- ifelse(input$type == "4 Year Undergraduate", 3, 
                       ifelse(input$type == "2 Year Undergraduate", 2, 1))
  
  
  tomapshiny <- tomap %>% 
    filter(tomap[income] <= input$max) %>% 
    filter(SCH_DEG == schooltype) %>% 
    filter(ACTna <= input$act) %>% 
    filter(SATna <= input$sat)
  

text <- c("Average Net Tuition for:", "under $30,000:", "$30,000-$48,000:", "$48,000-$75,000:", "$75,000-$110,000:", "above $110,000:")

  
  content <- paste(sep = "<br/>",
  as.character(tomapshiny$INSTNM),
  "ACT 75th Percentile:", as.character(tomapshiny$ACT75),
  "Average SAT Score:", as.character(tomapshiny$SAT_AVG),
  text[1], 
  text[2], as.character(tomapshiny$NPT41), 
  text[3], as.character(tomapshiny$NPT42), 
  text[4], as.character(tomapshiny$NPT43), 
  text[5], as.character(tomapshiny$NPT44),
  text[6], as.character(tomapshiny$NPT45))
  
  
  leaflet(data = tomapshiny) %>% 
    addProviderTiles("CartoDB.Positron") %>% 
    addCircles(radius = 3000, color = "navy", fillOpacity=0.4, 
               popup = ~content)
  
})

```

Data Sources:

* College Scorecard <https://collegescorecard.ed.gov/data/>
* National Association of College and Universtiy Buisness Officers <http://www.nacubo.org/Research/NACUBO-Commonfund_Study_of_Endowments/Public_NCSE_Tables/Total_Market_Value_of_Endowments.html>
* National Center for Education Statistics <https://nces.ed.gov/ipeds/datacenter/InstitutionByName.aspx>

