---
title: "College Affordability"
author: "Alden Cowap"
date: "11/30/2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, cache = TRUE}

## Get data and clean data

setwd("/Users/aldencowap/Documents/Fall 2016/Data Science/Final_Project")
library(dplyr)
library(pdftools)
library(stringr)
library(tm)
library(tidyr) 
library(SnowballC)
library(tidyverse)
library(stringdist)
library(USAboundaries)
library(maptools)
library(sp)
library(broom)
library(leaflet)
library(RColorBrewer)
library(stargazer)

datavar <- read.csv("variables.csv")
#data2015 <- read.csv("CollegeScorecard_Raw_Data/MERGED2014_15_PP.csv")
#data2014 <- read.csv("CollegeScorecard_Raw_Data/MERGED2013_14_PP.csv")
data2013 <- read.csv("CollegeScorecard_Raw_Data/MERGED2012_13_PP.csv")
needblind <- read.csv("needblindschools.csv")

endowment <- read.csv("endowment2013.csv")

endowment2 <- endowment %>% mutate(stringlabel = as.character(Rank)) %>% 
  filter(!(Rank %in% Rank[1:12])) %>% 
  filter(!(Rank %in% Rank[42:51])) %>% 
  mutate(first = substr(stringlabel, 1, 1)) %>% 
  filter(first %in% c(0:9))

states <- as.vector(state.abb)

endowmentname <- endowment2 %>% 
  mutate(stringlabel = str_replace_all(stringlabel, "<cd>", ".")) %>% 
  mutate(namebf = removeNumbers(stringlabel)) %>% 
  mutate(namebf = removePunctuation(namebf)) %>% 
  mutate(name = removeWords(namebf, states)) 

uniname <- as.vector(endowmentname[["namebf"]])


endowmentnums <- endowmentname %>% 
  mutate(nums = str_replace_all(stringlabel,"[^0-9 , .]","")) %>% 
  mutate(nums2 = str_replace_all(stripWhitespace(nums), " ", "/")) %>% 
  mutate(nums2 = str_replace_all(nums2, "[. ,]", "")) %>% 
  mutate(nums2 = str_replace_all(nums2, "///", "/")) %>% 
  mutate(nums2 = str_replace_all(nums2, "//", "/")) %>% 
  separate(nums2, into = c("rank", "endow2013", "endow2012", "change"), sep = "/") %>% 
  mutate(name = str_replace_all(name, " ", "")) %>% 
  mutate(name = str_replace(name, "CarnegieMellonUniversityv", "CarnegieMellonUniversity")) %>% 
  mutate(name = str_replace(name, "GeorgetownUniversityDC", "GeorgetownUniversity")) %>% 
  mutate(name = str_replace(name, "LawrenceUniversityofWisconsin", "LawrenceUniversity")) %>% 
  mutate(name = str_replace(name, "NorthCarolinaATStateUniversity", "NorthCarolinaStateUniversity")) %>% 
  mutate(name = str_replace(name, "PennsylvaniaStateUniversityii", "PennsylvaniaStateUniversity")) %>% 
  mutate(name = str_replace(name, "SanJoseStateUniversityTowerFoundation", "SanJoseStateUniversity")) %>% 
  mutate(name = str_replace(name, "StJohnsCollegeix", "StJohnsCollege")) %>% 
  mutate(name = str_replace(name, "UniversityofIllinoisFoundationiv", "UniversityofIllinoisatChicago")) %>% 
  mutate(name = str_replace(name, "FloridaStateUniversityFoundation", "FloridaStateUniversity")) %>% 
  mutate(name = str_replace(name, "UniversityofNorthCarolinaatChapelHillFoundations", "UniversityofNorthCarolina")) 



data2013temp <- data2013 %>% mutate(name = str_replace_all(as.character(INSTNM), " ", "")) %>% 
  mutate(name = str_replace(name, "ColumbiaUniversityintheCityofNewYork", "ColumbiaUniversity")) %>% 
  mutate(name = str_replace(name,"CooperUnionfortheAdvancementofScienceandArt", "CooperUnion")) %>% 
  mutate(name = str_replace(name,"FairleighDickinsonUniversity-MetropolitanCampus", "FairleighDickinsonUniversity")) %>%   
  mutate(name = str_replace(name,"HolyCrossCollege", "HolyCross")) %>%   
  mutate(name = str_replace(name,"JewishTheologicalSeminaryofAmerica", "JewishTheologicalSeminary")) %>%  
  mutate(name = str_replace(name,"Lewis&ClarkCollege", "LewisClarkCollege")) %>%   
  mutate(name = str_replace(name,"MountSaintMaryCollege", "MountSaintMarysCollege")) %>% 
  mutate(name = str_replace(name,"NorthCarolinaA&TStateUniversity", "NorthCarolinaStateUniversity")) %>%  
  mutate(name = str_replace(name,"PennsylvaniaStateUniversity-MainCampus", "PennsylvaniaStateUniversity")) %>%   
  mutate(name = str_replace(name,"StJohn'sCollege", "StJohnsCollege")) %>%   
  mutate(name = str_replace(name,"TulaneUniversityofLouisiana", "TulaneUniversity")) %>%   
  mutate(name = str_replace(name,"UniversityofNewHampshire-MainCampus", "UniversityofNewHampshire")) %>%   
  mutate(name = str_replace(name,"UniversityofNorthCarolinaatChapelHill", "UniversityofNorthCarolina")) %>%   
  mutate(name = str_replace(name,"UniversityofVirginia-MainCampus", "UniversityofVirginia")) %>%   
  mutate(name = str_replace(name,"UniversityofWashington-SeattleCampus", "UniversityofWashington"))

data2013temp <- data2013temp %>% 
  mutate(NPT41_PRIV = as.numeric(as.character(NPT41_PRIV))) %>%
  mutate(NPT42_PRIV = as.numeric(as.character(NPT42_PRIV))) %>% 
  mutate(NPT43_PRIV = as.numeric(as.character(NPT43_PRIV))) %>% 
  mutate(NPT44_PRIV = as.numeric(as.character(NPT44_PRIV))) %>% 
  mutate(NPT45_PRIV = as.numeric(as.character(NPT45_PRIV))) %>% 
  mutate(NPT41_PUB = as.numeric(as.character(NPT41_PUB))) %>% 
  mutate(NPT42_PUB = as.numeric(as.character(NPT42_PUB))) %>% 
  mutate(NPT43_PUB = as.numeric(as.character(NPT43_PUB))) %>% 
  mutate(NPT44_PUB = as.numeric(as.character(NPT44_PUB))) %>% 
  mutate(NPT45_PUB = as.numeric(as.character(NPT45_PUB))) %>% 
  mutate(NPT41 = ifelse(is.na(NPT41_PRIV), NPT41_PUB, NPT41_PRIV)) %>% 
  mutate(NPT42 = ifelse(is.na(NPT42_PRIV), NPT42_PUB, NPT42_PRIV)) %>% 
  mutate(NPT43 = ifelse(is.na(NPT43_PRIV), NPT43_PUB, NPT43_PRIV)) %>% 
  mutate(NPT44 = ifelse(is.na(NPT44_PRIV), NPT44_PUB, NPT44_PRIV)) %>% 
  mutate(NPT45 = ifelse(is.na(NPT45_PRIV), NPT45_PUB, NPT45_PRIV)) %>% 
  mutate(NUM4 = as.numeric(as.character(NUM4_PRIV))) %>% 
  mutate(NUM4 = ifelse(is.na(NUM4), as.numeric(as.character(NUM4_PUB)), NUM4)) %>% 
  mutate(SCH_DEG = as.numeric(as.character(SCH_DEG)))


total2013 <- inner_join(data2013temp, endowmentnums, by = "name") %>% arrange(rank)


latlong <- read_csv("latlong.csv") %>% 
  mutate(Name = Institution) %>% 
  mutate(Name = str_replace_all(Name, " ", "")) %>% 
  mutate(Name = str_replace_all(Name, "[. , ']", "")) %>% 
  rename(Ziplat = ZIP)
  
sat <- read_csv("sat.csv") %>% select(-Institution)

char <- inner_join(latlong, sat, by="UnitID")

total2013_lat <- inner_join(total2013, char, by=c("UNITID" = "UnitID"))

allschools_lat <- inner_join(data2013temp, char, by=c("UNITID" = "UnitID"))

needblind <- needblind %>% mutate(school = str_replace_all(Name.of.School, "[^a-z A-Z]", "")) %>% 
  mutate(school = str_replace_all(school, " ", "")) %>% 
  mutate(school = str_replace(school, "MassachusettsInstituteofTechnologyMIT", "MassachusettsInstituteofTechnology")) %>% 
  mutate(school = str_replace(school, "NewYorkUniversityNYU", "NewYorkUniversity")) %>% 
  mutate(school = str_replace(school, "UniversityofSouthernCaliforniaUSC", "UniversityofSouthernCalifornia")) %>% 
  mutate(school = str_replace(school, "WashingtonUniversity", "WashingtonUniversityinStLouis")) %>% 
  mutate(school = str_replace(school, "HarvardCollege", "HarvardUniversity")) %>% 
  mutate(school = str_replace(school, "NorthCarolinaStateUniversityNCSU", "NorthCarolinaStateUniversity")) %>% 
  mutate(school = str_replace(school,"PennState", "PennsylvaniaStateUniversity")) %>% 
  mutate(school = str_replace(school,"SanJoseStateUniversitySJSU", "SanJoseStateUniversity"))
  

blindnames <- as.vector(needblind$school)


blinddata <- total2013_lat %>% filter(name %in% blindnames)

name <- as.vector(blinddata$name)

antidata <- needblind %>% filter(!(school %in% name))

tograph <- blinddata %>% 
  mutate(endow = as.numeric(endow2013)) %>% 
  mutate(avg_cost = as.numeric(as.character(INC_PCT_M1))) %>% 
  filter(!is.na(avg_cost))

text <- c("Average Net Tuition for:", "under $30,000:", "$30,000-$48,000:", "$48,000-$75,000:", "$75,000-$110,000:", "above $110,000:")

TEMP <- tograph %>% select(endow, INSTNM, p_w_grant, NPT41, NPT42, NPT43, NPT44, NPT45)

TEMP <- TEMP %>% gather(income, amount, 4:8)

TEMP <- TEMP %>% 
  mutate(income = ifelse(income=="NPT41", text[2], 
                         ifelse(income=="NPT42", text[3], 
                                ifelse(income == "NPT43", text[4], 
                                       ifelse(income == "NPT44", text[5], text[6])))))


TEMP_no_nc <- TEMP %>% filter(endow!=32959)

total2013_lat <- total2013_lat %>% mutate(endow = as.numeric(as.character(endow2013)))

TEMP2 <- total2013_lat %>% select(endow, INSTNM, p_w_grant, NPT41, NPT42, NPT43, NPT44, NPT45)

TEMP2 <- TEMP2 %>% gather(income, amount, 4:8)

TEMP2 <- TEMP2 %>% 
    mutate(income = ifelse(income=="NPT41", text[2], 
                         ifelse(income=="NPT42", text[3], 
                                ifelse(income == "NPT43", text[4], 
                                       ifelse(income == "NPT44", text[5], text[6])))))

```

## Overview

1. In a previous statistics paper my group showed that amongst the top 50 liberal arts colleges endowment was correlated with:
  + average grant
  + number of students on financial aid
2. College is notoriously expensive
3. The listed tuition doesn't take into consideration fianancial aid

  
## The Data

* **Combination of 3 data sets:**
1. College Scorecard
2. National Association of College and Universtiy Buisness Officers
3. National Center for Education Statistics

## Endowment vs. Net Tuition

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(stringdist)
library(USAboundaries)
library(maptools)
library(sp)
library(broom)
library(leaflet)
library(RColorBrewer)
library(stargazer)
ggplot(data=TEMP2, aes(x = log10(endow), y = amount, color = income)) + 
  geom_smooth(method="lm", se = TRUE) + 
  labs(title = "All Schools", x = "Log of Endowment", y = "Net Tuition")
```

## Endwoment vs. Net Tuition for Need-Blind Schools

```{r, echo = FALSE}
ggplot(data=TEMP_no_nc, aes(x = log10(endow), y = amount, color = income)) + 
  geom_smooth(method="lm", se = TRUE) + 
  geom_point() + 
  labs(title = "Need Blind Schools", x = "Log of Endowment", y = "Net Tuition")
```

## Endwoment vs. Percentage on Financial Aid

```{r, echo = FALSE}
ggplot(data = TEMP, aes(x=log10(endow), y=p_w_grant)) + 
  geom_smooth(method = "lm") + 
  geom_point() + 
  labs(title = "Need Blind Schools", x="Log of Endowment", y = "Percent of Students on Financial Aid")
```

## Regressions

```{r star, results = 'asis', warning=FALSE, message=FALSE}
fitlowest <- lm(NPT41 ~ log(endow) + p_w_grant, data=tograph)
fitsecond <- lm(NPT42 ~ log(endow) + p_w_grant, data=tograph)
fitthird <- lm(NPT43 ~ log(endow) + p_w_grant, data=tograph)
fitfourth <- lm(NPT44 ~ log(endow) + p_w_grant, data=tograph)
fithigh <- lm(NPT45 ~ log(endow) + p_w_grant, data=tograph)

stargazer(fitlowest, fitsecond, fithigh, title="Regression Results", align=TRUE, type = "html",
          covariate.labels=c("Log of Endowment","Percent on Aid"), 
          dep.var.labels = c("under $30","$30-48", "above $110") ,
          keep.stat="n")

```

## Leaflet Interactive Map

* While this is all interesting, how do you make this information helpful?